A Bioconductor Workflow for Processing and Analysing Spatial Proteomics Data

### 2016-07-08 bioc-workflow.Rmd
Added more on TL

## The use-case: prediction sub-cellular localisation in pluripotent embryonic mouse stem cells

@@

These, as well as many other data are directly available as properly
structured and annotated computational object from the 
`r Biocexptpkg("pRolocdata")` experiment package. In this workflow, we
will start with a description of how to generate these ad hoc objects
starting from any arbitrary spreadsheets, as produced by many popular
third-party applications.

@@

Missing data are a recurrent issue in mass spectrometry applications,
and should be addressed independently of this workflow
[@Webb-Robertson:2015;@Lazar:2016]. In [@Gatto:2014b], we have
described how a high content in missing values in spatial proteomics
data and their inappropriate handling leads to a reduction of
sub-cellular resolution. Missing data can be imputated using

`r Biocpkg("MSnbase")`'s `impute` function. The method underlying the
imputation method is then determined by a `methods` parameter. In our
particular case, missing values are indicative of protein groups that
were not profiles in both replicates.

@@

In addition to high quality MS-based quantitative proteomics data there exist a number of other sources of information that are freely available in the public domain that may be useful to assign a protein to its sub-cellular niche. For example, imaging from immunofluorescence microscopy, protein
annotations and sequences, and protein-protein interactions among others, represent a rich and vast source of complementary information. We can integrate this auxiliary information with our primary MS-based quantitative data using a paradigm known as transfer learning (TL). The integration of data between different technologies is one of the biggest challenges in computational biology to date and the `r Biocpkg("pRoloc")` package provides functionality to do such analyses. We recently developed two transfer learning algorithms using a $k$-NN and SVM framework and applied them to the task of protein localisation prediction [@Breckels:2016].

@@

Using the functions `setAnnotationParams` and `makeGoSet` we can contruct an auxiliary `MSnSet` of GO terms, from the primary data's features i.e. the protein accession numbers. All the GO terms associated to each accession number are retrieved and used to create a binary matrix where a 1 (0) at position (i; j) indicates that term j has (not) been used to annotate protein i. The GO terms are retrieved from an appropriate repository using the `r Biocpkg("biomaRt")` package. The specific `r Biocpkg("Biomart")` repository and query will depend on the species under study and the type of identifiers. The first step is to condtruct the annotation parameters that will enable to perform the query, this is done using `setAnnotationParams`. Typing into the R console `par <- setAnnotationParams()` will present two menus, firstly asking you to identify the species of study, and then what type of identifier is used to annotate the proteins in your `MSnSet`. It is also possible to pass patterns to match the species e.g. in the code chunk below we pass "Mus musculus", and the identifier used e.g. "Uniprot/Swissprot", for the `Biocpkg("Biomart")` query.

```{r setap}
par <- setAnnotationParams(inputs = c("Mus musculus", 
                                      "UniProt/Swissprot"))

@@

```{r makego}
gocc <- makeGoSet(hl, params = par, 
                  namespace = "cellular_component", 
                  evidence = NULL)
gocc
```

@@

The *k*-NN TL algorithm estimates optimal weights for the different data sources and the spatial niches described for the data at hand with the `knntlOptimisation` function.

>>>>

## The use-case: predicting sub-cellular localisation in pluripotent embryonic mouse stem cells

@@

These, as well as many other data are directly available as properly
structured and annotated computational object from the 
`r Biocexptpkg("pRolocdata")` experiment package. In this workflow, we
will start with a description of how to generate these ad hoc objects
starting from an arbitrary spreadsheets, as produced by many popular
third-party applications.

@@

Missing data are a recurrent issue in mass spectrometry applications,
and should be addressed independently of this workflow
[@Webb-Robertson:2015;@Lazar:2016]. In [@Gatto:2014b], we have
described how a high content in missing values in spatial proteomics
data and their inappropriate handling leads to a reduction of
sub-cellular resolution. Missing data can be imputated using 
`r Biocpkg("MSnbase")`'s `impute` function. The method underlying the
imputation method is then determined by a `methods` parameter. In our
particular case, missing values are indicative of protein groups that
were not profiles in both replicates.

@@

In addition to high quality MS-based quantitative proteomics data there exist a number of other sources of information that are freely available in the public domain that may be useful to assign a protein to its sub-cellular niche. For example, imaging from immunofluorescence microscopy, protein
annotations and sequences, and protein-protein interactions among others, represent a rich and vast source of complementary information. We can integrate this auxiliary information with our primary MS-based quantitative data using a paradigm known as transfer learning (TL). The integration of data between different technologies is one of the biggest challenges in computational biology to date and the `r Biocpkg("pRoloc")` package provides functionality to do such analyses. We recently developed two transfer learning algorithms using a *k*-NN and SVM framework and applied them to the task of protein localisation prediction [@Breckels:2016]. In this section we will begin with explaining the concept of transfer learning and then show how to apply this in the frame of spatial proteomics and protein localisation prediction.

@@

Using the functions `setAnnotationParams` and `makeGoSet` we can contruct an auxiliary `MSnSet` of GO terms, from the primary data's features i.e. the protein accession numbers. All the GO terms associated to each accession number are retrieved and used to create a binary matrix where a 1 (0) at position (i; j) indicates that term j has (not) been used to annotate protein i. The GO terms are retrieved from an appropriate repository using the `r Biocpkg("biomaRt")` package. The specific `r Biocpkg("Biomart")` repository and query will depend on the species under study and the type of identifiers. The first step is to construct the annotation parameters that will enable to perform the query, this is done using `setAnnotationParams`. Typing into the R console `par <- setAnnotationParams()` will present two menus, firstly asking you to identify the species of study, and then what type of identifier you have used to annotate the proteins in your `MSnSet`. It is also possible to pass patterns to match the species e.g. in the code chunk below we pass "Mus musculus", and the identifier type for our data (see `featureNames(hl)`) which is "Uniprot/Swissprot", for the `Biocpkg("Biomart")` query.

```{r setap, eval=FALSE}
par <- setAnnotationParams(inputs = c("Mus musculus", 
                                      "UniProt/Swissprot"))

@@

```{r makego, eval=FALSE}
gocc <- makeGoSet(hl, params = par, 
                  namespace = "cellular_component", 
                  evidence = NULL)
gocc
```

@@

We begin by performing the optimisation. The `knntlOptimisation` function estimates the optimal weights for the different data sources and the spatial niches described for the data.

- first need to do k-NN optimisation to get best k's for P and A
- explain weight matrix of thetas, `by` and `length,out`
- explain how to do this on the cluster, in practise too computationally heavy on a single processor

```{r tlopt, eval=FALSE}

getMarkers(hl, fcol = "phenoDisco.Input")

opt <- knntlOptimisation(hl, gocc,
                         fcol = "phenoDisco.Input",
                         by = .5, 
                         times = 1,
                         xval = 5, k = c(3,3),
                         BPPARAM = par
  )
```
